---
title: "SEIR Model Analysis"
format: 
  html: 
    df-print: paged
    embed-resources: true
---
```{r, message=FALSE}
library(outbreaks)
library(tidyverse)
library(rstan)
library(gridExtra)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```

```{r}
glimpse(influenza_england_1978_school)
```

```{r}
# time series of cases
cases <- influenza_england_1978_school$in_bed  # Number of students in bed

# total count
N <- 763

# times
n_days <- length(cases)
t <- seq(0, n_days, by = 1)
t0 = 0 
t <- t[-1]

# initial conditions
i0 <- 1
e0 <- 2  # Assuming some exposed individuals initially
s0 <- N - i0 - e0
r0 <- 0
y0 = c(S = s0, E = e0, I = i0, R = r0)

# data for Stan
data_seir <- list(n_days = n_days, y0 = y0, t0 = t0, ts = t, N = N, cases = cases)
data_seir_new_priors <- data_seir
data_seir_old_priors <- data_seir

# number of MCMC steps
niter <- 2000

```


```{r}
model <- stan_model("SEIR_Schoo 2.stan")
```

```{r}
# Fit model with new informative priors
data_seir_new_priors$use_new_priors <- 1
fit_seir_new <- sampling(model, data = data_seir_new_priors, iter = niter, chains = 4, seed = 0)

# Fit model with old priors
data_seir_old_priors$use_new_priors <- 0
fit_seir_old <- sampling(model, data = data_seir_old_priors, iter = niter, chains = 4, seed = 0)


```

```{r}
pars <- c('beta', 'sigma', 'gamma', "R0", "incubation_period", "recovery_time")

# Print summaries
print(fit_seir_new, pars = pars)
print(fit_seir_old, pars = pars)

```

```{r}
# Plot density for new priors
p1 <- stan_dens(fit_seir_new, pars = pars, separate_chains = TRUE) + ggtitle("Parameter Densities with New Priors")
# Plot density for old priors
p2 <- stan_dens(fit_seir_old, pars = pars, separate_chains = TRUE) + ggtitle("Parameter Densities with Old Priors")

grid.arrange(p1, p2, ncol = 1)

```

Parameter Density Plots (New Priors): This plot shows the posterior distributions of the parameters  estimated from the SEIR model using the new priors. The density curves for each parameter represent the range of values that the parameters can take, based on the data and the new priors.
Parameter Density Plots (Old Priors): This plot shows the same posterior distributions but using the old priors.

```{r}
# Predictions for new priors
smr_pred_new <- cbind(as.data.frame(summary(
  fit_seir_new, pars = "pred_cases", probs = c(0.05, 0.5, 0.95))$summary), t, cases)
colnames(smr_pred_new) <- make.names(colnames(smr_pred_new)) # to remove % in the col names

# Predictions for old priors
smr_pred_old <- cbind(as.data.frame(summary(
  fit_seir_old, pars = "pred_cases", probs = c(0.05, 0.5, 0.95))$summary), t, cases)
colnames(smr_pred_old) <- make.names(colnames(smr_pred_old)) # to remove % in the col names

# Plot predictions for new priors
p3 <- ggplot(smr_pred_new, mapping = aes(x = t)) +
  geom_ribbon(aes(ymin = X5., ymax = X95.), fill = 'red', alpha = 0.35) +
  geom_line(mapping = aes(x = t, y = X50.), color = 'red') + 
  geom_point(mapping = aes(y = cases)) +
  labs(x = "Day", y = "Number of students in bed") +
  ggtitle("Predicted Cases vs. Actual Cases (New Priors)")

# Plot predictions for old priors
p4 <- ggplot(smr_pred_old, mapping = aes(x = t)) +
  geom_ribbon(aes(ymin = X5., ymax = X95.), fill = 'red', alpha = 0.35) +
  geom_line(mapping = aes(x = t, y = X50.), color = 'red') + 
  geom_point(mapping = aes(y = cases)) +
  labs(x = "Day", y = "Number of students in bed") +
  ggtitle("Predicted Cases vs. Actual Cases (Old Priors)")

p3

p4

```
Predicted Cases vs. Actual Cases (New Priors): This plot compares the predicted number of cases (students in bed) from the SEIR model with the new priors against the actual observed cases. The red ribbon represents the 90% credible interval, the red line shows the median prediction, and the black points are the actual cases.
Predicted Cases vs. Actual Cases (Old Priors): Similarly, this plot compares the predictions using the old priors against the actual observed cases.

```{r}
params <- lapply(t, function(i){sprintf("y[%s,3]", i)}) # number of infected for each day

# Infected predictions for new priors
smr_y_new <- as.data.frame(summary(fit_seir_new, pars = params, probs = c(0.05, 0.5, 0.95))$summary)
colnames(smr_y_new) <- make.names(colnames(smr_y_new)) # to remove % in the col names

# Infected predictions for old priors
smr_y_old <- as.data.frame(summary(fit_seir_old, pars = params, probs = c(0.05, 0.5, 0.95))$summary)
colnames(smr_y_old) <- make.names(colnames(smr_y_old)) # to remove % in the col names

# Plot infected for new priors
p5 <- ggplot(smr_y_new, mapping = aes(x = t)) +
  geom_ribbon(aes(ymin = X5., ymax = X95.), fill = 'red', alpha = 0.35) +
  geom_line(mapping = aes(x = t, y = X50.), color = 'red') + 
  labs(x = "Day", y = "Number of infected students") +
  ggtitle("1 Predicted Number of Infected Students (New Priors)")

# Plot infected for old priors
p6 <- ggplot(smr_y_old, mapping = aes(x = t)) +
  geom_ribbon(aes(ymin = X5., ymax = X95.), fill = 'red', alpha = 0.35) +
  geom_line(mapping = aes(x = t, y = X50.), color = 'red') + 
  labs(x = "Day", y = "Number of infected students") +
  ggtitle(" 2 Predicted Number of Infected Students (Old Priors)")

p5
p6

```
Predicted Number of Infected Students (New Priors): This plot shows the predicted number of infected students over time, according to the SEIR model with the new priors. The red ribbon represents the 90% credible interval, and the red line shows the median prediction.
Predicted Number of Infected Students (Old Priors): This plot shows the same predictions using the old priors.
